---
output:
  pdf_document: default
  html_document: default
---
# Hierarchical Model, Mixture Distribution（階層モデル、混合分布）

2026-4-22

## 復習{-}

- 条件付き確率（密度）関数、周辺確率（密度）関数
- Bayesian Rule, Bayes estimators

## 階層モデル（pp.163~、現代数理統計の基礎 pp.65）


- 例： Binomial-Poisson Hierarchy

  - 分布と期待値

### 定理１：Conditional expectation（証明）

## Mixture Distributionの定義

- 例： Normal scale mixture distribution (現 pp.65)

- 例： Binomial-Poisson Hierarchy

  - 分布と期待値

- 例： Poisson-Gamma distribution (課題 $\rightarrow$ 4.32)

- 例： Noncentral $\chi^2$ distribution

- 例： Beta-binomial Hierarchy

### 定理２：Conditional variance identity（証明）

- 例： Beta-binomial Hierarchy

### EM algorithm (pp.327、現代数理統計の基礎 pp.263)

- 例： Multiple Poisson rates
- 定理*
- EM vs MLE*
- Missing data issue*
  
- Kullback–Leibler (KL) divergence: $KL(Q||P)$ - a measure of how much an approximating probability distribution $Q$ is different from a true probability distribution $P$ (現代数理統計の基礎 pp.195)

- Newton-Raphson method (現代数理統計の基礎 pp.262)

## コメント{-}

- Posterior 分布の導出
- 不確実性を階層的にモデル化

## 応用{-}

- GLMM: generalized linear mixed effects model（一般化線形混合モデル）

  - その特例： GLM
  - その一般化： Bayesian hierarchical model、階層ベイズ
  - Meta-analysis　データの統合
  
- Bayes analysis (pp.371)

  - Empirical Bayes analysis
  - Hierarchical Bayes analysis 
  - Gibbs sampling

- EM 
  
  - 混合分布の推定　(現代数理統計の基礎 pp.264 例11.4)
  - 分類モデル

## R experiment

Suppose $X\sim N(\mu, 1)$, where $\mu$ is unknown.

The observable data are $\{0,2,x\}$, where $x$ indicates the missing value. 

We can use the following EM algorithm to find $\hat\mu$

```{r}
mu <- 2   ## Initial value
tol <- 1e-6  ## Difference between mu_new and mu
diff <- Inf  ## set some very large value as initial difference

while (diff > tol) { ## If difference is SMALLER than some very small value, then stop
  mu_new <- (0 + 2 + mu) / 3
  diff <- abs(mu_new - mu)
  mu <- mu_new
  print(mu)
}

```

Because:

1. Loglikelihood of $\{0,2,x\}$: $log L(\mu|x)\propto -\dfrac{1}{2}\{(0-\mu)^2+(2-\mu)^2+(x-\mu)^2\}$, where we omit all the constants.

2. $E_{X|\mu'}[log L(\mu|X)]\\\propto E_{X|\mu'}[-\dfrac{1}{2}\{(0-\mu)^2+(2-\mu)^2+(X-\mu)^2\}]\\=-\dfrac{1}{2}\{(0-\mu)^2+(2-\mu)^2+E_{X|\mu'}[(X-\mu)^2]\}\\=-\dfrac{1}{2}\{(0-\mu)^2+(2-\mu)^2+E_{X|\mu'}[X^2-2X\mu+\mu^2]\}\\=-\dfrac{1}{2}\{(0-\mu)^2+(2-\mu)^2+E_{X|\mu'}[X^2]-2E_{X|\mu'}[X]\mu+\mu^2]\}\\=-\dfrac{1}{2}\{(0-\mu)^2+(2-\mu)^2+(1+\mu'^2)-2\mu'\mu+\mu^2]\}$

3. Maximum $E_{X|\mu'}[log L(\mu|X)] \Rightarrow \mu = \dfrac{0+2+\mu'}{3}$ 

4. Set an initial value for $\mu'$ and derive $\mu$ until $\mu=\mu'$


### Gaussian Mixture Model

Suppose we observed some scores from 2 groups (e.g. $N(\mu_1, \sigma_1^2), N(\mu_2, \sigma_2^2)$). However we could not observe the group variable.

Data look like this

```{r}
y <- c(-0.39, 0.12, 0.94, 1.67, 1.76, 2.44, 3.72, 4.28, 4.92, 5.53, 
       0.06, 0.48, 1.01, 1.68, 1.80, 3.25, 4.12, 4.60, 5.28, 6.22)

hist(y) ## this is the histogram (empirical distribution) of data
```

Suppose $Z_i$ is the latent group variable (unobservable), where $P(Z_i=0)=1-\pi, P(Z_1=1)=\pi$. 

$Y_i|Z_i=0\sim N(\mu_1, \sigma_1^2)=\phi(y| \mu_1,\sigma_1); Y_i|Z_i=1\sim N(\mu_2, \sigma_2^2)=\phi(y| \mu_2,\sigma_2)$

Then the marginal pdf of $Y$: $f_Y(y)=(1-\pi)\phi(y| \mu_1,\sigma_1)+\pi\phi(y| \mu_2,\sigma_2)$

Likelihood function is 

$L(\pi,\mu_1,\sigma_1,\mu_2,\sigma_2) = \prod_{i=1}^N (1-\pi)\phi(\mu_1,\sigma_1|y_i)+\pi\phi(\mu_2,\sigma_2|y_i)$

We could use MLE to estimate the parameters $(\pi,\mu_1,\sigma_1,\mu_2,\sigma_2)$ but the derivation is complicated. 


Consider the EM algorithm.

```{r}
## EM
par <- c(0.5, 0.5, 1, 0.5, 1)  # Initial values of (pi, mu1, mu2, sigma1, sigma2)
tol <- 1e-6 
diff <- Inf  ## set some very large value as initial difference

while (diff > tol) { ## If difference is SMALLER than some very small value, then stop
  
  pi <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  sigma1 <- par[4]
  sigma2 <- par[5]
  
  ## ---- E-step ----
  f1 <- dnorm(y, mu1, sigma1) ## density of normal distribution
  f2 <- dnorm(y, mu2, sigma2)
  
  gamma <- pi * f2 / ((1 - pi) * f1 + pi * f2)
  
  ## ---- M-step ----
  pi_new <- mean(gamma)
  
  mu1_new <- sum((1 - gamma) * y) / sum(1 - gamma)
  mu2_new <- sum(gamma * y) / sum(gamma)
  
  sigma1_new <- sqrt(sum((1 - gamma) * (y - mu1_new)^2) / sum(1 - gamma))
  sigma2_new <- sqrt(sum(gamma * (y - mu2_new)^2) / sum(gamma))
  
  par_new <- c(pi_new, mu1_new, mu2_new, sigma1_new, sigma2_new)
  
  diff <- max(abs(par_new - par))
  par <- par_new

  print(par)
}
```

The detailed derivations of the parameters can be found online.

Continue with the model.

With the estimates, we have the marginal pdf of $Y$: $f_Y(y)=(1-0.4454)\phi(y| 1.0832,0.9008)+\pi\phi(y| 4.6559, 0.9049)$. (Attention: this is just the density, not the probability)

What we can do next is to predict the probabilities of a new point falling into each group.

Suppose a new point is observed as $y_{new}=3$, we can derive $f_Y(y_{new})$

```{r}
y <- 3
par ## the estimates
pi <- par[1]
mu1 <- par[2]
mu2 <- par[3]
sigma1 <- par[4]
sigma2 <- par[5]

f1 <- dnorm(y, mu1, sigma1)
f2 <- dnorm(y, mu2, sigma2)
  
fy <- (1-pi) * f1 + pi * f2
fy ## but this is not interesting
```
The results we care about is the classification

$P(Z=0 \mid y_{\text{new}}) 
= \dfrac{(1-\hat{\pi})\,\phi(y_{\text{new}} \mid \hat{\mu}_1,\hat{\sigma}_1)}
{f_Y(y_{\text{new}})}$

$P(Z=1 \mid y_{\text{new}}) 
= \dfrac{\hat{\pi}\,\phi(y_{\text{new}} \mid \hat{\mu}_2,\hat{\sigma}_2)}
{f_Y(y_{\text{new}})}
$



```{r}
## Predict the classification results

pz0 <- (1-pi) * f1/fy
pz1 <- pi * f2/fy
```

The probabilities of being classified into groups are shown as

```{r}
pz0
pz1
```

