# Hierarchical Model, Mixture Distribution（階層モデル、混合分布）

2026-4-22

## 復習{-}

- 条件付き確率（密度）関数、周辺確率（密度）関数
- Bayesian Rule, Bayes estimators

## 階層モデル（pp.163~、現代数理統計の基礎 pp.65）


- 例： Binomial-Poisson Hierarchy

  - 分布と期待値

### 定理１：Conditional expectation（証明）

## Mixture Distributionの定義

- 例： Normal scale mixture distribution (現 pp.65)

- 例： Binomial-Poisson Hierarchy

  - 分布と期待値

- 例： Poisson-Gamma distribution (課題 $\rightarrow$ 4.32)

- 例： Noncentral $\chi^2$ distribution

- 例： Beta-binomial Hierarchy

### 定理２：Conditional variance identity（証明）

- 例： Beta-binomial Hierarchy

### EM algorithm (pp.327、現代数理統計の基礎 pp.263)

- 例： Multiple Poisson rates
- 定理*
- EM vs MLE*
- Missing data issue*
  
- Kullback–Leibler (KL) divergence: $KL(Q||P)$ - a measure of how much an approximating probability distribution $Q$ is different from a true probability distribution $P$ (現代数理統計の基礎 pp.195)

- Newton-Raphson method (現代数理統計の基礎 pp.262)

## コメント{-}

- Posterior 分布の導出
- 不確実性を階層的にモデル化

## 応用{-}

- GLMM: generalized linear mixed effects model（一般化線形混合モデル）

  - その特例： GLM
  - その一般化： Bayesian hierarchical model、階層ベイズ
  - Meta-analysis　データの統合
  
- Bayes analysis (pp.371)

  - Empirical Bayes analysis
  - Hierarchical Bayes analysis 
  - Gibbs sampling

- EM 
  
  - 混合分布の推定　(現代数理統計の基礎 pp.264 例11.4)
  - 分類モデル

## R Experiment

Suppose $X\sim N(\mu, 1)$, where $\mu$ is unknown.

Observable data is $\{0,2,*\}$

We can use the following EM algorithm to find $\hat\mu$

```{r}
mu <- 2
tol <- 1e-6
diff <- Inf

while (diff > tol) {
  mu_new <- (0 + 2 + mu) / 3
  diff <- abs(mu_new - mu)
  mu <- mu_new
  print(mu_new)
}

```

It is the simple version of the following process. 

```{r}
mu <- 2
tol <- 1e-6
diff <- Inf
while (diff > tol) {
  
  EMl = function(mu2) integrate(
    {function(x) (dnorm(0, mu2, 1, log = TRUE)+
      dnorm(2, mu2, 1, log = TRUE)+
      dnorm(x, mu2, 1, log = TRUE)) *
        dnorm(x, mu, 1)}, -Inf, Inf)$value
  
  mu_new = optimize(EMl, c(-100, 100), maximum = TRUE)$maximum
  diff <- abs(mu_new - mu)
  mu <- mu_new
  print(mu_new)
}

```

Consider another example.

Suppose we observed some score from 2 groups (e.g. $N(\mu_1, \sigma_1^2), N(\mu_2, \sigma_2^2)$). However we could not observe the group variable.

Data look like this

```{r}
y <- c(-0.39, 0.12, 0.94, 1.67, 1.76, 2.44, 3.72, 4.28, 4.92, 5.53, 
       0.06, 0.48, 1.01, 1.68, 1.80, 3.25, 4.12, 4.60, 5.28, 6.22)

hist(y)
```

Suppose $Z_i$ is the latent group variable (unobservable), where $P(Z_i=0)=1-\pi, P(Z_1=1)=1-\pi$. 

$Y_i|Z_i=0\sim N(\mu_1, \sigma_1^2); Y_i|Z_i=1\sim N(\mu_2, \sigma_2^2)$

Then the marginal pdf of $Y$: $f_Y(y)=(1-\pi)\phi(y| \mu_1,\sigma_1)+\pi\phi(y| \mu_2,\sigma_2)$

Likelihood function is 

$L(\pi,\mu_1,\sigma_1,\mu_2,\sigma_2) = \prod_{i=1}^N (1-\pi)\phi(\mu_1,\sigma_1|y_i)+\pi\phi(\mu_2,\sigma_2|y_i)$

We could use MLE (numerical optimization) to estimate the parameters
```{r}
## MLE
ll <- function(par){
  
  pi <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  sigma1 <- par[4]
  sigma2 <- par[5]
  
  f1 <- dnorm(y, mu1, sigma1)
  f2 <- dnorm(y, mu2, sigma2)
  
  -sum(log((1 - pi)*f1 + pi*f2))
}

optim(c(0.5, 0.5, 1, 0.5, 1), ## initial value
      ll,method="L-BFGS-B",
      lower = c(1e-4, -Inf, -Inf, 1e-4, 1e-4),
      upper = c(1 - 1e-4, Inf, Inf, Inf, Inf))
```


EM algorithm 

```{r}
## EM
par <- c(0.5, 0.5, 1, 0.5, 1)  # (pi, mu1, mu2, sigma1, sigma2)
tol <- 1e-6
diff <- Inf

while (diff > tol) {
  
  pi <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  sigma1 <- par[4]
  sigma2 <- par[5]
  
  ## ---- E-step ----
  f1 <- dnorm(y, mu1, sigma1)
  f2 <- dnorm(y, mu2, sigma2)
  
  gamma <- pi * f2 / ((1 - pi) * f1 + pi * f2)
  
  ## ---- M-step ----
  pi_new <- mean(gamma)
  
  mu1_new <- sum((1 - gamma) * y) / sum(1 - gamma)
  mu2_new <- sum(gamma * y) / sum(gamma)
  
  sigma1_new <- sqrt(sum((1 - gamma) * (y - mu1_new)^2) / sum(1 - gamma))
  sigma2_new <- sqrt(sum(gamma * (y - mu2_new)^2) / sum(gamma))
  
  par_new <- c(pi_new, mu1_new, mu2_new, sigma1_new, sigma2_new)
  
  diff <- max(abs(par_new - par))
  par <- par_new
  # names(par) <- c("pi_new", "mu1_new", "mu2_new", "sigma1_new", "sigma2_new")
  
  print(par)
}
```

